Devito for large scale elastic modelling and anisotropic inversion

### 2020-04-20 paper.tex
Elastic section update.

While the subsurface image in the previous Section is obtained with
anisotropic propagators capable of mimicking the real physics, in order
to model both the kinematics and the amplitudes correctly elastic
propagators are required. These propagators are for example extremely
important for global seismology, as the Shear surface waves (component
ignored in TTI) are the most hazardous ones. In this section, we exploit
the tensor algebra language introduced in
\href{https://github.com/devitocodes/devito}{Devito} v4.0 to express
elastic waves in a compact and elegant notation.

The elastic isotropic wave-equation, parametrized by the Lamé parameters
$\lambda, \mu$ and the density $\rho$ reads:
%
\begin{equation}
\begin{aligned}
&\frac{1}{\rho}\frac{dv}{dt} = \nabla . \tau \\
&\frac{d \tau}{dt} = \lambda \mathrm{tr}(\nabla v) \mathbf{I}  + \mu (\nabla v + (\nabla v)^T)
\end{aligned}
\label{elas1}
\end{equation}
%
 where $v$ is a vector valued function with one component per cartesian
direction:
%
\begin{equation}
\begin{split}
v =  \begin{bmatrix} v_x(t, x, y) \\ v_y(t, x, y)) \end{bmatrix}
\end{split}
\label{partvel}
\end{equation}
%
 and the stress $\tau$ is a symmetric second-order tensor-valued
function:
%
\begin{equation}
\begin{aligned}
    \tau = \begin{bmatrix}\tau_{xx}(t, x, y) & \tau_{xy}(t, x, y)\\\tau_{xy}t, x, y) & \tau_{yy}(t, x, y)\end{bmatrix}.
\end{aligned}
\label{stress}
\end{equation}
%
 The discretization of such a set of coupled PDEs requires five
equations in two dimensions (two equations for the particle velocity and
three for stress) and nine equations in three dimensions (three particle
velocities and six stress equations). However, the mathematical
definition only requires two coupled vector/tensor-valued equations for
any number of dimensions.

@@

We have augmented the
\href{https://github.com/devitocodes/devito}{Devito} language with
tensorial objects to enable a straightforward -- and at the same time
mathematically rigorous -- definition of high-dimensional PDEs, such as
the elastic wave-equation in Equation~\ref{elas1}. This effort was
inspired by \cite{ufl}, a functional language for finite element
methods.

The extended \href{https://github.com/devitocodes/devito}{Devito}
language introduces two new types, \texttt{VectorFunction} (and
\texttt{VectorTimeFunction}) for vectorial objects such as the particle
velocity, and \texttt{TensorFunction} (and \texttt{TensorTimeFunction})
for second-order tensor objects (matrices) such as the stress. These new
objects are constructed the exact same way as the scalar
\texttt{Function} objects. They also automatically implement staggered
grid and staggered finite-differences with the possibility of half-node
averaging. Each component of a tensorial object -- a (scalar)
\href{https://github.com/devitocodes/devito}{Devito} \texttt{Function}
-- is accessible via conventional vector notation (i.e.
\texttt{v{[}0{]},\phantom{\ }t{[}0,1{]},....}).

With this extended language, the elastic wave-equation can be defined in
only four lines as follows:

@@

The \texttt{SymPy} expressions created by these commands can be
displayed with \texttt{sympy.pprint} as shown in
Figure~\ref{PrettyElas}. This representation reflects perfectly the
mathematics while still providing computational portability and
efficiency through the
\href{https://github.com/devitocodes/devito}{Devito} compiler. The
complete generated code for the elastic wave-equation with and without
MPI is made available at
\href{https://github.com/mloubout/SC20Paper/tree/master/gencode}{CodeSample}
in \texttt{elastic-so8.c} and \texttt{elastic-so8-mpi.c}.

\begin{figure*}
\centering
\includegraphics[width=1.000\hsize]{./Figures/vel_symb.png}
\caption{Update stencil for the particle velocity. The stencil for
updating the stress component is left out for readability, as the
equation does not fit onto a single page. However, it can be found in
the \href{https://github.com/devitocodes/devito}{Devito} tutorial on
elastic modelling on github.}\label{PrettyElas}
\end{figure*}

@@

We show the elastic particle velocity and stress for a broadly
recognized 2D synthetic model, the elastic
Marmousi-ii\cite{versteeg927, marmouelas} model. The wavefields are
shown on Figure~\ref{ElasWf} and its corresponding elastic shot records
are displayed in Figure~\ref{ElasShot}. Thes two figures show that the
wavefield is as expected purely acoustic in the water layer
(\texttt{t\_\{xy\}=0}) and transition perfectly at the ocean bottom to
an elastic wavefield. We can also clearly see the Shear wave wavefront
in the subsurface (around 1km depth). Thes two plots show that we
effectively solve the elastic wave-equation with a high-level
mathematical definition of the PDE that abstracts away all the
computational technicalities such as staggered grid analysis. The shot
records in Figure~\ref{ElasShot} also match the original data generated
by the creator of the elastic model in \cite{marmouelas}.

\begin{figure*}
\centering
\includegraphics[width=1.000\hsize]{./Figures/marmou_snap.png}
\caption{Particle velocities and stress at time $t=3\text{s}$ for a
source at 10m depth and \texttt{x=5\textbackslash{}text\{km\}} in the
marmousi-ii model.}\label{ElasWf}
\end{figure*}

\begin{figure*}
\centering
\captionsetup[subfigure]{labelformat=empty}
\subfloat[]{\includegraphics[width=0.300\hsize]{./Figures/pressure_marmou.png}}
\subfloat[]{\includegraphics[width=0.300\hsize]{./Figures/vz_marmou.png}}
\subfloat[]{\includegraphics[width=0.300\hsize]{./Figures/vx_marmou.png}}
\caption{Seismic shot record for 5sec of modelling. \texttt{a} is the
pressure (trace of stress tensor) at the surface (5m depth), \texttt{b}
is the vertical particle velocity and \texttt{c} is the horizontal
particle velocity at the ocean bottom (450m depth).}\label{ElasShot}
\end{figure*}

@@

Finally, we modelled three dimensional elastic data in the Cloud to
demonstrate the scalability of
\href{https://github.com/devitocodes/devito}{Devito} to cluster-size
problems in the Cloud. The model we chose mimics the reference model in
geophysics known as the SEAM model \cite{fehler2011seam} that is a
three dimensional extreme-scale synthetic representation of the
subsurface. The physical dimension of the model are
\texttt{45kmx35kmx15km} then discretized with a grid spacing of
\texttt{20mx20mx10m} that led to a computational grid of
\texttt{2250x1750x1500} grid points (5.9 billion grid points). One of
the main challenges of elastic modelling is the extreme memory cost due
to the number of wavefield. For a three dimensional propagator, a
minimum of 21 fields need to be stored:

@@

These 21 fields, with the grid we just describe, leads to a minimum of
461Gb of memory for modelling only. For this experiment, we obtained
access to small HPC VM on azure called \texttt{Standard\_H16r} that are
16 cores Intel Xeon E5 2667 v3 with no hyperthreading and used 32 nodes
for a single source experiment (we solved a single wave-equation). We
used a 12th order discretization that leads to 2.8TFLOP/time-step to be
computed for this model and propagated the elastic wave for 16 seconds
(23000 time steps). The modelling ran in 16 hours which converts to
1.1TFLOP/s. While these number may seem low, the elastic kernel is
extremely memory bound, while the TTI kernel is nearly compute bound
(see rooflines in \cite{louboutin2016ppf, devito-api, devito-compiler})
making it more computationally efficient, in particular in combination
with MPI. Future work with involve working on the InfiniBand enabled and
true HPC VM on azure to achieve Cloud performance on par with state of
the art Cluster performance. From the performance obtained in this
experiment, and assuming a fairly standard setup of 5000 independent
source experiments, computing a synthetic dataset would require 322
EFLOP to be computed (23k time-steps x 2.8TFLOP/time-step x 500
sources), or using the full scalability and computing all sources in
parallel would lead to 5.5PFLOP/s from our single source experiment
result.

>>>>

While the subsurface image obtained in section \ref{d-imaging-example-on-azure}
utilized anisotropic propagators capable of mimicking intricate physics, in order
to model both the wave kinematics and amplitudes correctly, elastic
propagators are required. These propagators are, for example, extremely
important in global seismology since shear surface waves (which are
ignored in acoustic models) are the most hazardous. In this section, we exploit
the tensor algebra language introduced in
\href{https://github.com/devitocodes/devito}{Devito} v4.0 to express
an elastic model with compact and elegant notation.

The isotropic elastic wave-equation, parametrized by the so-called Lamé parameters
$\lambda, \mu$ and the density $\rho$ reads:
%
\begin{equation}
\begin{aligned}
&\frac{1}{\rho}\frac{dv}{dt} = \nabla . \tau \\
&\frac{d \tau}{dt} = \lambda \mathrm{tr}(\nabla v) \mathbf{I}  + \mu (\nabla v + (\nabla v)^T)
\end{aligned}
\label{elas1}
\end{equation}
%
 where $v$ is a vector valued function with one component per cartesian
direction:
%
\begin{equation}
\begin{split}
v =  \begin{bmatrix} v_x(t, x, y) \\ v_y(t, x, y)) \end{bmatrix}
\end{split}
\label{partvel}
\end{equation}
%
 and the stress $\tau$ is a symmetric second-order tensor-valued
function:
%
\begin{equation}
\begin{aligned}
    \tau = \begin{bmatrix}\tau_{xx}(t, x, y) & \tau_{xy}(t, x, y)\\\tau_{xy}t, x, y) & \tau_{yy}(t, x, y)\end{bmatrix}.
\end{aligned}
\label{stress}
\end{equation}
%
 The discretization of \eqref{elas1} and \eqref{stress} requires five
equations in two dimensions (two equations for the particle velocity and
three for the stress) and nine equations in three dimensions (three for velocity
and six for stress). However, the mathematical
definition only require two coupled vector/tensor-valued equations for
any number of dimensions.

@@

We have augmented the
\href{https://github.com/devitocodes/devito}{Devito} language with
tensorial objects to enable straightforward and
mathematically rigorous definitions of high-dimensional PDEs, such as
the elastic wave-equation defined in \eqref{elas1}. This implementation was
inspired by \cite{ufl}, a functional language for finite element
methods.

The extended \href{https://github.com/devitocodes/devito}{Devito}
language introduces two new types, \texttt{VectorFunction}/
\texttt{VectorTimeFunction} for vectorial objects such as the particle
velocity, and \texttt{TensorFunction}/\texttt{TensorTimeFunction}
for second-order tensor objects (matrices) such as the stress. These new
objects are constructed in the same manner as scalar
\texttt{Function} objects. They also automatically implement staggered
grid and staggered finite-differences with the possibility of half-node
averaging. Each component of a tensorial object -- a (scalar)
\href{https://github.com/devitocodes/devito}{Devito} \texttt{Function}
-- is accessible via conventional vector notation (i.e.
\texttt{v{[}0{]},\phantom{\ }t{[}0,1{]},....}).

With this extended language, the elastic wave-equation defined in
\eqref{elas1} and \eqref{stress} can be expressed in only four lines
of code:

@@

The \texttt{SymPy} expressions created by these commands can be
displayed with \texttt{sympy.pprint} as shown in
Figure~\ref{PrettyElas}. This representation reflects perfectly the
mathematics while maintaining computational portability and
efficiency through the
\href{https://github.com/devitocodes/devito}{Devito} compiler. The
complete generated code for the elastic wave-equation with and without
MPI is made available at
\href{https://github.com/mloubout/SC20Paper/tree/master/gencode}{CodeSample}
in \texttt{elastic-so8.c} and \texttt{elastic-so8-mpi.c}.

\begin{figure*}
\centering
\includegraphics[width=1.000\hsize]{./Figures/vel_symb.png}
\caption{Particle velocity update stencil. The stress update
is omitted for readability (the
equation does not fit into a single page). However, it can be found in
the \href{https://github.com/devitocodes/devito}{Devito} tutorial on
elastic modelling on github.}\label{PrettyElas}
\end{figure*}

@@

To demonstrate the efficacy of the elastic implementation outlined
above we utilized a broadly recognized 2D synthetic model, the elastic
Marmousi-ii\cite{versteeg927, marmouelas} model. The wavefields are
shown on Figure~\ref{ElasWf} and its corresponding elastic shot records
are displayed in Figure~\ref{ElasShot}. These two figures show that the
wavefield is, as expected, purely acoustic in the water layer
(\texttt{t\_\{xy\}=0}) and transitions correctly at the ocean bottom to
an elastic wavefield. We can also clearly see the shear wave-front
in the subsurface (at a depth of approximately 1km). Figures \ref{ElasWf}
and \ref{ElasShot} demonstrate that this high-level
\href{https://github.com/devitocodes/devito}{Devito} implementation
of the elastic wave-equation is effective and accurate.
Importantly, in constructing this model within the
\href{https://github.com/devitocodes/devito}{Devito} DSL framework,
computational technicalities such as the staggered grid analysis are
abstracted away. We note that the shot records displayed in Figure~\ref{ElasShot}
match the original data generated by the creator of this elastic model
(\cite{marmouelas}).

\begin{figure*}
\centering
\includegraphics[width=1.000\hsize]{./Figures/marmou_snap.png}
\caption{Particle velocities and stress at time $t=3\text{s}$ for a
10m deep source and \texttt{x=5\textbackslash{}text\{km\}} in the
marmousi-ii model.}\label{ElasWf}
\end{figure*}

\begin{figure*}
\centering
\captionsetup[subfigure]{labelformat=empty}
\subfloat[]{\includegraphics[width=0.300\hsize]{./Figures/pressure_marmou.png}}
\subfloat[]{\includegraphics[width=0.300\hsize]{./Figures/vz_marmou.png}}
\subfloat[]{\includegraphics[width=0.300\hsize]{./Figures/vx_marmou.png}}
\caption{Seismic shot record for 5sec of modelling. \texttt{a} is the
pressure (trace of stress tensor) at the surface (5m depth), \texttt{b}
and \texttt{c} display, respectively, the vertical and horizontal
particle velocity at the ocean bottom (450m depth).}\label{ElasShot}
\end{figure*}

@@

Finally, three dimensional elastic data was modelled in the Cloud to
demonstrate the scalability of
\href{https://github.com/devitocodes/devito}{Devito} to cluster-size
problems. The model used in this experiment mimics a reference model in
geophysics known as the SEAM model \cite{fehler2011seam}, a
three dimensional extreme-scale synthetic representation of a
subsurface. The physical dimensions of the model are
\texttt{45kmx35kmx15km} discretized with a grid spacing of
\texttt{20mx20mx10m} leading to a computational grid of
\texttt{2250x1750x1500} grid points (5.9 billion grid points). One of
the main challenges of elastic modelling is the extreme memory cost owing
to the number of wavefields (a minimum of 21 fields in a three dimensional
propagator) that need to be stored:

@@

These 21 fields, for the 5.9 billion point grid defined above, lead to a
minimum memory requirement of 461Gb for modelling alone. For this experiment,
access was obtained for small HPC VMs (on azure) called \texttt{Standard\_H16r}.
These VMs contain 16 core Intel Xeon E5 2667 v3 chips, with no hyperthreading,
and 32 nodes were used for a single source experiment
(i.e. a single wave-equation was solved). We
used a 12th order discretization in space that led to 2.8TFLOP/time-step being
computed by this model and the elastic wave was propagated for 16 seconds
(23000 time steps). Completion of this modelling run took 16 hours, converting to
1.1TFLOP/s. While these numbers may appear low, it should be noted that
the elastic kernel is extremely memory bound, while the TTI kernel is nearly
compute bound (see rooflines in \cite{louboutin2016ppf, devito-api, devito-compiler})
making it more computationally efficient, particularly in combination
with MPI. Future work will involve working on InfiniBand enabled and
true HPC VMs on azure to achieve Cloud performance on par with that of state of
the art HPC clusters. Extrapolating from the performance obtained in this
experiment, and assuming a fairly standard setup of 5000 independent
source experiments, computing an elastic synthetic dataset would require 322
EFLOPs (23k time-steps x 2.8TFLOP/time-step x 500
sources), or utilizing the full scalability and computing all sources in
parallel this becomes 5.5PFLOP/s.
